Bandit Multiclass Linear Classification: Efficient Algorithms for the Separable Case

### 2019-01-22 ICML-submission/notions-of-linear-separability.tex
Update notions-of-linear-separability.tex
Changed some frac to slash
add experiments

We define two notions of linear separability for multiclass classification. The
first notion is the standard notion of linear separability used in the proof of
the mistake bound for the \textsc{Multiclass Perceptron} algorithm. The second
notion is stronger, i.e. more restrictive. However, it is more suitable for the
bandit setting, since it allows for a simple and efficient algorithm; see
Section~\ref{section:algorithm-for-strongly-linearly-separable-data}.

\begin{definition}
[Linear separability]
\label{definition:weak-linear-separability}
Let $(V,\ip{\cdot}{\cdot})$ be an inner product space, $K$ be a positive
integer, and $\gamma$ be a positive real number. 
We say that labeled examples $(x_1, y_1),
(x_2, y_2), \dots, (x_T, y_T) \in V \times \{1,2,\dots,K\}$ are 
%said to be
\begin{enumerate}
\item
\emph{weakly linearly separable with a margin $\gamma$} if there exist vectors
$w_1, w_2, \dots, w_K \in V$ such that
\begin{align}
\label{equation:weak-linear-separability-1}
& \sum_{i=1}^K \norm{w_i}^2 \le 1 \; , \\
\label{equation:weak-linear-separability-2}
& \begin{gathered}
\forall t \in \{1,2,\dots,T\} \quad \forall i \in \{1,2,\dots, K\} \setminus \{y_t\} \\
\ip{x_t}{w_{y_t}} \ge \ip{x_t}{w_i} + \gamma \; ,
\end{gathered}
\end{align}
\item
\emph{strongly linearly separable with a margin $\gamma$} if there exist vectors
$w_1, w_2, \dots, w_K \in V$ such that
\begin{align}
\label{equation:strong-linear-separability-1}
& \sum_{i=1}^K \norm{w_i}^2 \le 1 \; , \\
\label{equation:strong-linear-separability-2}
& \forall t \in \{1,2,\dots,T\} \qquad \qquad \ip{x_t}{w_{y_t}} \ge \frac \gamma 2 \; , \\
\label{equation:strong-linear-separability-3}
& \begin{gathered}
\forall t \in \{1,2,\dots,T\} \quad \forall i \in \{1,2,\dots, K\} \setminus \{y_t\} \\
\ip{x_t}{w_i} \le - \frac \gamma 2 \; .
\end{gathered}
\end{align}
\end{enumerate}
\end{definition}

The notion of weak linear separability is a standard assumption used in the
full-information setting to upper bound the number of mistakes of the
\textsc{Multiclass Perceptron} algorithm.
%%%%%
(Recall our discussion on the topic in Section~\ref{section:introduction}.)
%Namely, \citet{Crammer-Singer-2003}
%prove that if the examples are weakly linearly separable with a margin $\gamma$
%and the norm of the examples is bounded by $R$, then \textsc{Multiclass
%Perceptron} algorithm makes at most $\left\lfloor 2(R/\gamma)^2 \right \rfloor$
%mistakes. It is a folklore result that \textsc{Multiclass Perceptron} is
%essentially optimal in the sense that any (possibly randomized) algorithm must
%make $\frac{1}{2} \left\lfloor (R/\gamma)^2 \right \rfloor$ mistakes in the
%worst case. For completeness, we state the \textsc{Multiclass Perceptron}
%algorithm, the upper and lower bound on the number of mistakes, and their proofs
%in Appendix~\ref{section:multiclass-perceptron-proofs} in the supplementary
%material.
%%%%%

@@

\begin{figure}
\begin{center}
\input{figures/weakly-linearly-separable-examples-with-margin}
\end{center}
\caption[]{A set of labeled examples in $\R^2$. The examples belong to
$K=3$ classes colored white, gray and black respectively. Each class lies in a
$120^\circ$ wedge. In other words, each class lies in an intersection of two
halfspaces. While the examples are weakly linearly separable with a positive margin
$\gamma$, they are \emph{not} strongly linearly separable with any positive
margin $\gamma$. For instance, there does \emph{not} exist a linear separator
that separates the examples belonging to the gray class from the examples
belonging to the remaining two classes.
}
\label{figure:weakly-linearly-separable-examples-with-margin}
\end{figure}

>>>>

We define two notions of linear separability for multiclass classification. The
first notion is the standard notion of linear separability used in the proof of
the mistake bound for the \textsc{Multiclass Perceptron} algorithm. The second
notion is stronger, i.e. more restrictive. %However, it is more suitable for the
%bandit setting, since it allows for a simple and efficient algorithm; see
%Section~\ref{section:algorithm-for-strongly-linearly-separable-data}.

\begin{definition}
[Linear separability]
\label{definition:weak-linear-separability}
Let $(V,\ip{\cdot}{\cdot})$ be an inner product space, $K$ be a positive
integer, and $\gamma$ be a positive real number. 
We say that labeled examples $(x_1, y_1),
(x_2, y_2), \dots, (x_T, y_T) \in V \times \{1,2,\dots,K\}$ are 
%said to be
\begin{enumerate}
\item
\emph{weakly linearly separable with a margin $\gamma$} if there exist vectors
$w_1, w_2, \dots, w_K \in V$ such that
\begin{align}
\label{equation:weak-linear-separability-1}
& \sum_{i=1}^K \norm{w_i}^2 \le 1 \; , \\
\label{equation:weak-linear-separability-2}
& \begin{gathered}
\forall t \in \{1,2,\dots,T\} \quad \forall i \in \{1,2,\dots, K\} \setminus \{y_t\} \\
\ip{x_t}{w_{y_t}} \ge \ip{x_t}{w_i} + \gamma \; ,
\end{gathered}
\end{align}
\item
\emph{strongly linearly separable with a margin $\gamma$} if there exist vectors
$w_1, w_2, \dots, w_K \in V$ such that
\begin{align}
\label{equation:strong-linear-separability-1}
& \sum_{i=1}^K \norm{w_i}^2 \le 1 \; , \\
\label{equation:strong-linear-separability-2}
& \forall t \in \{1,2,\dots,T\} \qquad \qquad \ip{x_t}{w_{y_t}} \ge \gamma/2 \; , \\
\label{equation:strong-linear-separability-3}
& \begin{gathered}
\forall t \in \{1,2,\dots,T\} \quad \forall i \in \{1,2,\dots, K\} \setminus \{y_t\} \\
\ip{x_t}{w_i} \le - \gamma/2 \; .
\end{gathered}
\end{align}
\end{enumerate}
\end{definition}

%The notion of weak linear separability is a standard assumption used in the
%full-information setting to upper bound the number of mistakes of the
%\textsc{Multiclass Perceptron} algorithm.
%%%%%
%(Recall our discussion on the topic in Section~\ref{section:introduction}.)
%Namely, \citet{Crammer-Singer-2003}
%prove that if the examples are weakly linearly separable with a margin $\gamma$
%and the norm of the examples is bounded by $R$, then \textsc{Multiclass
%Perceptron} algorithm makes at most $\left\lfloor 2(R/\gamma)^2 \right \rfloor$
%mistakes. It is a folklore result that \textsc{Multiclass Perceptron} is
%essentially optimal in the sense that any (possibly randomized) algorithm must
%make $\frac{1}{2} \left\lfloor (R/\gamma)^2 \right \rfloor$ mistakes in the
%worst case. For completeness, we state the \textsc{Multiclass Perceptron}
%algorithm, the upper and lower bound on the number of mistakes, and their proofs
%in Appendix~\ref{section:multiclass-perceptron-proofs} in the supplementary
%material.
%%%%%

@@

\begin{figure}
\begin{center}
 \scalebox{.8}{\input{figures/weakly-linearly-separable-examples-with-margin}}
\end{center}
\caption[]{A set of labeled examples in $\R^2$. The examples belong to
$K=3$ classes colored white, gray and black respectively. Each class lies in a
$120^\circ$ wedge. In other words, each class lies in an intersection of two
halfspaces. While the examples are weakly linearly separable with a positive margin
$\gamma$, they are \emph{not} strongly linearly separable with any positive
margin $\gamma$. For instance, there does \emph{not} exist a linear separator
that separates the examples belonging to the gray class from the examples
belonging to the remaining two classes.
}
\label{figure:weakly-linearly-separable-examples-with-margin}
\end{figure}
